Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-17 paper2.qmd
section:appendix; process:write

@@

For instance, If researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as a variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion. Under these circumstances, researchers firt identify the *conditional average causal effect* (CACE, @eq-CACE)

@@

From the CACE, they identify the ACE from associational quantities as in @eq-mACE. Here, $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].

$$
\begin{aligned}
  \tau & = E[Y_{i} \cond do(T_{i}=1)] - E[Y_{i} \cond do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]
\end{aligned}
$$ {#eq-mACE}

@@

## SCMs and DAGs {#sec-appendix-C .appendix appendix-style=plain}

@@

<!-- 5. Continuing with the motivating example -->
Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B. SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example. In this context, the variable $X$ (socio-economic status of the school) acts as a confounder in the relationship between the teaching method $T$ and the outcome $Y$. The figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Additional, the figures show one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.

@@

Plausible causal structure the scenario outlined in @sec-appendix-B.
:::

## The probabilistic implications of SCMs and DAGs {#sec-appendix-D .appendix appendix-style=plain}

@@

<!-- 4. can you illustrate this concept? -->
The five fundamental building blocks described in @sec-appendix-C clearly illustrate which (in)dependencies can SMCs and DAGs predict. For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent or d-separated ($X_{1} \dsep X_{3}$). This conclusion follows from expressing their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$ [@Neal_2020, pp. 24]. Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$. Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendix-C and the CPDs.

@@

<!-- 5. Continuing with the motivating example -->
Researchers can revisit the scenario in @sec-appendix-B using these insights. By applying the BNF to the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) = P(Y \cond T, X) P(T \cond X) P(X)$. Assuming the *direct effect* of $T$ on $Y$ ($T \rightarrow Y$, black path) is absent, researchers can infer from the joint probability that $Y$ is not unconditionally independent of $T$ $(Y \dsep T)$. This dependency arises because a non-causal path exists between $T$ and $Y$ through $X$ ($T \leftarrow X \rightarrow Y$, gray path), indicating that $X$ is a confounder. However, researchers can observe that conditioning on $X$, depicted as a square in DAG [-@fig-mdag_example1a], results in $Y \dsep T \cond X$, eliminating the confounding effect and allowing the identification of the direct effect $T \rightarrow Y$ if it exists.

@@

::: {#fig-scm_example1 fig-align="center"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{T}(X,e_{T}) \\
  Y & := f_{Y}(t,X,e_{Y}) \\
  e_{T} & \dsep e_{X} \\
  e_{T} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

Plausible causal structure the scenario outlined in @sec-appendix-B.
:::

## From probability to causality {#sec-appendix-E .appendix appendix-style=plain}

<!-- 1. But, how do we identify causal effects?  -->
In addition to the assumptions required to encode probabilistic information, SCMs and DAGs rely on the modularity assumption to represent interventions $P(Y_{i} \cond do(T_{i}=t))$ (see @sec-appendix-B). Defining these interventions enables researchers to identify the ACE from associational quantities through statistical conditioning, effectively disentangling causation from mere association.

The modularity assumption underpins the concepts of Truncated Factorization and manipulated graphs, which are essential for representing interventions within SCMs and DAGs. The *modularity assumption* posits that intervening on a node modifies only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp. 34]. Building on this principle, the *Truncated Factorization* (TF) simulates physical interventions by removing specific functions from the conceptual model and substituting them with constants, preserving the rest of the structure unchanged [Pearl_2010]. The probabilistic implications of this factorization are formalized in @eq-trunc_factor, where $S$ denotes the subset of variables $X_{p}$ directly influenced by the intervention. Complementing the TF, *manipulated graphs* are DAGs where specific edges have been removed to reflect the effects of interventions [@Neal_2020, pp. 34].

@@

Using the TF, researchers can define the *backdoor adjustment*, a method for identifying causal effects. This adjustment specifies that if a variable $X_{p} \in S$ serves as a *sufficient adjustment set* for the effect of $X_{a}$ on $X_{b}$, then the ACE can be identified using @eq-backdoor [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014] (see @sec-appendix-B).

@@

<!-- 6. Continuing with the motivating example -->
<!-- Researchers can apply these new insights to revisit the scenario in @sec-appendix-B. From the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a], they can infer that the effect of $T$ on $Y$ is confounded by $X$.  -->

<!-- This is evident because the effect decomposes into a *direct causal effect* ($T \rightarrow Y$, black path) and a non-causal effect arising from the confounder $X$ ($T \leftarrow X \rightarrow Y$, gray path).  -->

<!-- Using @eq-net_factor and @eq-trunc_factor, researchers can show that the joint probability of the observed variables in these structures can be expressed as $P(Y, T, X) = P(Y \cond T, X) P(T \cond X) P(X)$. Assuming intervention, this expression becomes $P(Y, X \cond do(T=t)) = P(Y \cond T,X) P(X)$, and by marginalizing over $X$, they obtain $P(Y \cond do(T=t)) = \sum_{X} P(Y \cond T,X) P(X)$ [@Neal_2020, pp. 36]. -->

<!-- Assuming the causal structure is correct, researchers can identify the CACE of $T$ on $Y$ by conditioning on $X$ and then marginalize this effect over $X$ to obtain the ACE. Conditioning on $X$ means restricting the focus to schools with a specific socio-economic status $X$ [@Neal_2020, pp. 32]. Notably, researchers often depict conditioning as a square around a variable. Furthermore, the conditioning process leads to the manipulated DAG [-@fig-mdag_example1b], equivalent to the conditioned DAG [-@fig-mdag_example1a]. -->

<!--  From these distributions, the ACE can be expressed as $\tau = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]$ (see @sec-appendix-B). -->

@@

::: {#fig-scm_example1 fig-align="center"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := t \\
  Y & := f_{Y}(t,X,e_{Y}) \\
  e_{T} & \dsep e_{X} \\
  e_{T} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

Plausible causal structure the scenario outlined in @sec-appendix-B.
:::

## The estimation process {#sec-appendix-F .appendix appendix-style=plain}

<!-- 1. How do we use Bayesian inference? -->
Researchers can use Bayesian inference methods to estimate the ACE. The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \cond \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity. After observing empirical data, researchers can update the priors to posterior distributions using Bayes' rule [@Jeffreys_1998]:

@@

<!-- 3. How do DAGs can help? -->
Nevertheless, researchers can manage the complexity of the likelihood by assuming that variables exhibit specific local (in)dependencies. These assumptions enhance model tractability and facilitate the estimation process. As described in @sec-appendix-D, researchers can use SCMs and DAGs to formalize these local (in)dependencies, which leads to the BNF of the likelihood (@eq-like_BNF). From this point, any probabilistic programming language can model this simpler structure and generate the parameter's posterior distribution from its priors using @eq-bayes_rule.

>>>>

### The general idea {#sec-appendix-B1 .appendix appendix-style=plain}

@@

For instance, If researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as a variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion. Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)

@@

From the CACE, they identify the ACE from associational quantities as in @eq-mACE1. This identification process is commonly known as the *backdoor adjustment*. Here, $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].

$$
\begin{aligned}
  \tau & = E[Y_{i} \cond do(T_{i}=1)] - E[Y_{i} \cond do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]
\end{aligned}
$$ {#eq-mACE1}

@@

### Diving into the specifics {#sec-appendix-B2 .appendix appendix-style=plain}

#### SCMs and DAGs {#sec-appendix-B21 .appendix appendix-style=plain}

@@

<!-- 5. Continuing with the motivating example -->
Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B1. SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example. In this context, the variable $X$ (socio-economic status of the school) acts as a confounder in the relationship between the teaching method $T$ and the outcome $Y$. The figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Additional, the figures show one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.

@@

Plausible causal structure the scenario outlined in @sec-appendix-B1.
:::

#### The probabilistic implications of SCMs and DAGs {#sec-appendix-B22 .appendix appendix-style=plain}

@@

<!-- 4. can you illustrate this concept? -->
The five fundamental building blocks described in @sec-appendix-B21 clearly illustrate which (in)dependencies can SMCs and DAGs predict. For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent or d-separated ($X_{1} \dsep X_{3}$). This conclusion follows from expressing their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$ [@Neal_2020, pp. 24]. Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$. Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendix-B21 and the CPDs.

@@

<!-- 5. Continuing with the motivating example -->
Researchers can revisit the scenario in @sec-appendix-B1 using these insights. By applying the BNF to the SCM [-@fig-scm_example1a] and DAG [-@fig-mdag_example1a] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$. From this expression, researchers can infer that the outcome $Y$ is not unconditionally independent of the teaching method $T$ $(Y \dsep T)$. This dependency arises from two key structures: a direct causal path from the teaching method $T$ to the outcome $Y$, represented by the two-connected-nodes structure $T \rightarrow Y$ (black path), and a confounding non-causal path from the teaching method $T$ to the outcome $Y$ through the socio-economic status of the school $X$, represented by the fork structure $T \leftarrow X \rightarrow Y$ (gray path).

These probabilistic insights also indicate that narrowing the focus to a subset of the population defined by the socio-economic status of the school $X$, achieved by conditioning on $X$ and depicted by a square in DAG [-@fig-mdag_example1a], effectively blocks all confounding non-causal path introduced by the fork structure. Nevertheless, since $Y$ remains not d-separated from $T$ even after conditioning $(Y \ndsep T \cond X)$, this lack of d-separation can only be attributed to the direct causal effect $T \rightarrow Y$.

@@

::: {#fig-scm_example1a fig-align="center"}
$$
\begin{aligned}
  X & := x \\
  T & := f_{T}(x,e_{T}) \\
  Y & := f_{Y}(T,x,e_{Y}) \\
  e_{T} & \dsep e_{X} \\
  e_{T} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

Plausible causal structure the scenario outlined in @sec-appendix-B1.
:::

#### From probability to causality {#sec-appendix-B23 .appendix appendix-style=plain}

<!-- 1. Transforming probabilistic association into causation?  -->
The structural approach to causal inference translate the probabilistic insights from @sec-appendix-B23 into actionable strategies for identifying the ACE from associational quantities. The approach achieve this by relying on the modularity assumption and defining interventions of the form $P(Y_{i} \cond do(T_{i}=t))$ (see @sec-appendix-B1). The *modularity assumption* posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp. 34].

<!-- 2. additional necessary concepts  -->
The modularity assumption underpins the concepts of manipulated graphs and Truncated Factorization, which are essential for representing interventions within SCMs and DAGs. *Manipulated graphs* simulate physical interventions by removing specific edges from a DAG, while preserving the remaining structure unchanged [@Neal_2020, pp. 34]. In parallel, *Truncated Factorization* (TF) achieves a similar simulation by removing specific functions from the conceptual model and replacing them with constants, while keeping the rest of the structure unchanged [Pearl_2010]. The probabilistic implications of this factorization are formalized in @eq-trunc_factor, where $S$ represents the subset of variables $X_{p}$ directly influenced by the intervention.

@@

<!-- 3. How to use these concepts to finally define the ACE -->
Using the TF, researchers can define the *backdoor adjustment* to identify the ACE. This adjustment states that if a variable $X_{p} \in S$ serves as a *sufficient adjustment set* for the effect of $X_{a}$ on $X_{b}$, then the ACE can be identified using @eq-backdoor [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. The sufficient adjustment set (potentially empty) must block all non-causal paths between $X_{a}$ and $X_{b}$ without introducing new paths. If such a set exists, then $X_{a}$ and $X_{b}$ are *d-separated* by $X_{p}$ ($X_{a} \dsep X_{b} \cond X_{p}$) [@Pearl_2009], and $X_{p}$ satisfies the *backdoor criterion* [@Neal_2020, pp. 37].

@@

<!-- 4. revisiting the motivating example -->
With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example1a] and DAG [-@fig-mdag_example1a] is correct. This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure. In this context, researchers can apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$. They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE, as expressed in @eq-mACE2 (see @sec-appendix-B). Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example1a] is equivalent to the manipulated DAG [-@fig-mdag_example1b], because $X$ satisfies the backdoor criterion.

$$
\begin{aligned}
  \tau & = E[Y_{i} \cond do(T_{i}=1)] - E[Y_{i} \cond do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]
\end{aligned}
$$ {#eq-mACE2}

@@

::: {#fig-scm_example1b fig-align="center"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := t \\
  Y & := f_{Y}(t,X,e_{Y}) \\
  e_{T} & \dsep e_{X} \\
  e_{T} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

Plausible causal structure the scenario outlined in @sec-appendix-B22.
:::

## The estimation process {#sec-appendix-C .appendix appendix-style=plain}

<!-- 1. How do we use Bayesian inference? -->
Ultimately, researchers can use Bayesian inference methods to estimate the ACE. The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \cond \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity. After observing empirical data, researchers can update the priors to posterior distributions using Bayes' rule [@Jeffreys_1998]:

@@

<!-- 3. How do DAGs can help? -->
Nevertheless, researchers can manage the complexity of the likelihood by assuming that variables exhibit specific local (in)dependencies. These assumptions enhance model tractability and facilitate the estimation process. As described in @sec-appendix-B22, researchers can use SCMs and DAGs to formalize these local (in)dependencies, which leads to the BNF of the likelihood (@eq-like_BNF). From this point, any probabilistic programming language can model this simpler structure and generate the parameter's posterior distribution from its priors using @eq-bayes_rule.
